An exposition on the propriety of restricted Boltzmann machines

### 2016-10-01 writing/draft.Rmd
body of instability rewrite complete

For RBMs, an analogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibles that differ at only one coordinate.

\begin{align*}
&\max\left\{\log\frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v^*_1, \dots, v^*_V)}: (v_1, \dots, v_V) \text{ and } (v^*_1, \dots, v^*_V) \in \{-1, 1\}^V \text{ differ by one component} \right\} \\
&\quad = \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right)}: v_k \not= v^*_k \right\} \\
& \quad \ge \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right)}: v_k \not= v^*_k \right\} \\
&\quad = \max\limits_{k \in 1, \dots, V}\left\{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\}\right. \\
& \quad \quad \left. - \max\limits_{\boldsymbol h \in \{-1,1\}^H}
\left\{\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right\} - H\log2: v_k \not= v^*_k \right\} \\
&\quad \ge \max\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - \min\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - H\log 2
\end{align*}

Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$. Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two observations that differ in only one element is greater than $C$. So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.

@@

```{r child='appendix.Rmd'}
```

>>>>

\begin{prop}
\label{prop:instab}
If $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$, then $\Delta_V(\boldsymbol \theta) > C$.
\end{prop}

In other words, if the quantity (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e. instability mentioned earlier).

Additionally, unstable RBM models are connected to degenerate models (placing all probability on a small pieace of the sample space). Define a model set 
$$
M_{\epsilon, \boldsymbol \theta} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$.

\begin{prop}
\label{prop:degen}
For an unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
\end{prop}

In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability. Proofs of both proposition \ref{prop:instab} and proposition \ref{prop:degen} are provided in appendix \ref{instability-proofs}. Propositions \ref{prop:instab}-\ref{prop:degen} have counterpart results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixednumber of parameters (i.e. our results are broader for a more complicated model form for the visibles). Essentially, with our generalized definition of *unstable* models, one can establish instability results as in @schweinberger2011instability under assumptions that the sample space is finite and all model outcomes have a non-zero probability.

To later numerically investigate unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
$$
\frac{1}{V}R(\boldsymbol \theta) \text{ for } R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2.
$$
That is, it holds that, in (\ref{eqn:elpr}),
\begin{align}
\frac{1}{V} \text{ELPR}_V(\boldsymbol \theta) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta). \label{eqn:lowerbnd}
\end{align}
(A proof of (\ref{eqn:lowerbnd}) appears in appendix \ref{instability-proofs}.)

The benefit of (\ref{eqn:lowerbnd}) is that $R(\boldsymbol \theta)/V$ is simple to compute (i.e. compared to (\ref{eqn:elpr}), $R(\boldsymbol \theta)$ requires no summation over hidden variables) and that propositions \ref{prop:instab}-\ref{prop:degen} would continue to hold upon replcing (\ref{eqn:elpr}) with $R(\boldsymbol \theta)/V$, due to (\ref{eqn:lowerbnd}). Hence, larger values of $R(\boldsymbol \theta)/V$ indicate unstable RBM models.

@@

```{r child='appendix-center.Rmd'}
```

## Instability proofs

```{r child='appendix-instability.Rmd'}
```
